---
title: "Juvenile Snowshoe Hares in Bonanza Creek Experimental Forest"
author: "By: Mikel Irigoyen"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(kableExtra)
library(effsize)
library(broom)
```


```{r}
hares <- read.csv(here("data", "bonanza_hares.csv")) #read it in!
```

### 1. Introduction

This report shows a conscientious overview of differences in juvenile hare weight within the Bonanza Creek Experimental Forest from 1998 - 2012 (Kielland et al. 1999). Specifically, differences between male and female hares and assessing if weight correlates to hindfoot length.

### 2. Data and Analyses

Data on trapped hares including site, weight, and hindfoot length was provided by Bonanza Creek Long Term Ecological Research Network (LTER), which is located in the boreal forest of interior Alaska. Over 300 juvenile hares were trapped over a span of 14 years (1998 - 2012). Following exploratory data visualization, weights of male and female hares are looked at using a two-sample t-test. Differences between groups are described by Cohen's *d* effect size. The relationship between weight and hindfoot length explored by simple linear regression. All analyses are in R version 4.0.2 using RStudio version 1.3.1073. 


### 3. Exploratory Findings


```{r, fig.keep='first'}
hares_by_year <- hares %>%  #create subset
  filter(age == "j") %>%  #filter by age: juvenile
  mutate(date_updated = mdy(date)) %>%  #use lubridate to change date format
  mutate(year_updated = year(date_updated)) %>%#create new column for year
  
  group_by(year_updated) %>% #group by year
  count(year_updated)   #use count function to find n per year

ggplot(hares_by_year,
       aes(x = year_updated, y = n))+
  geom_col(color = "black", fill = "steelblue") + # make a column plot showing hares caught per year
  
  scale_x_continuous(breaks = c(1998, 1999, 2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012)) +
  # maybe a better way to do this but it seems to work... just updating the axis to show all years
  
  labs(x = "",
       y = "Hares trapped",
       title = "Annual juvenile hare trap counts") + #adding labels
       theme_bw() #%>%  # new theme 
summarize(hares_by_year, average = mean(n, na.rm = TRUE)) # find the mean number of hairs trapped per year
```
**Figure.1** Amount of juvenile hares trapped annually between 1998 - 2012 at all sites. There is a large amount of variability in hares trapped year to year. With the highest number or trapped hares occurring in 1999 (126) and multiple years (1998, 2002, 2009) showing zero juvenile hares trapped. There may be variability in days spent trapping year to year. To remedy this inconsistency, it may prove prudent to standardize the amount of days spent trapping and time of year moving forward.







```{r}
hare_weights <- hares %>% #create subset 
  filter(age == "j") %>% #filter by age: juvenile
  select(weight, sex, grid) %>% #select by weight, sex, and site (grid)
  group_by(sex, grid) #group by sex and site 

ggplot(hare_weights, aes(sex, weight, fill = sex)) +  #create a boxplot
geom_boxplot(aes(sex, weight)) + 
facet_wrap(~grid) + #facet wrap to show all 3 sites side by side 
labs(caption = "Juvenile Weight and Sex Comparision by Site", x = "", y = "Weight (g)") + # captions and labels 
  stat_summary(fun.y=mean, geom = "point", shape = 22, size = 2) + # cool way to show mean on the boxplot
  theme_bw() + #new theme 
  theme(plot.caption = element_text(hjust = 0, size = 12), legend.title = element_blank()) + 
  #change location and size of plot caption
  scale_fill_discrete(labels=c("Female", "Male", "NA")) # trying to change label names on legend 

```

**Figure 2.** Juvenile hare weights by sex and site (grid). Red (female), teal (male), and gray (NA) boxes indicate sex at each site. Box endpoints indicate the 25th and 75th percentile values; the black box indicate the  mean value for each species, respectively. Data: (Kielland et al. 1999).
```{r}
hare_comparison <- hare_weights %>%  #create subset 
  
  na.omit(hare_comparison) %>%  # get rid of NA data so we can focus on just males and females
  group_by(sex) %>% # group by sex (m or f)
  summarize(mean_weight = mean(weight, na.rm = TRUE), #find the mean 
            sd_weight = sd(weight, na.rm = TRUE), #find the standard deviation
            sample_size = n()) #find the sample size n 

hare_comparison %>%  #dataframe to use for kableExtra
  kbl(caption = "Juvenile weight comparison", col.names = c("Sex", "Mean (g)", "SD (g)", "Sample Size")) %>%  #adding a caption for the table and changing column names
  kable_classic(full_width = F, html_font = "Cambria") #classic look, changed width of table, and font style 

```

**Table 1.** 
```{r}
ggplot(hare_weights, aes(weight)) + # creating a ggplot by weights 
  geom_histogram() + #histogram
  facet_wrap(~sex) #facet wrap by sex to see m and f side by side

ggplot(hare_weights, aes(sample = weight)) + # everything same as histogram except this time we re making a qqplot... dont forget the "sample =" for qq plots
  geom_qq() + #qq plot
  facet_wrap(~sex) #facet wrap by sex to split plot into m and f 

female_hares <- hare_weights %>% #create subset
  filter(sex == "f") %>%  #filter by sex (f)
  pull(weight) # get the vectors for the weight values

male_hares <- hare_weights %>% #do same as above but for males
  filter(sex == "m") %>% 
  pull(weight)

hare_ttest <- t.test(female_hares, male_hares) #t test time, female vs male

effect_size <- cohen.d(na.omit(female_hares), na.omit(male_hares)) #effect size
```

## Relationship between juvenile weight & hind foot length.

```{r}
hind_foot_hares <- hares %>% #create subset
  filter(age == "j")  # filter by juvenile age 
 
 ggplot(hind_foot_hares, aes(weight, hindft)) + #ggplot
  geom_point() + #point
  theme_light() + #new theme
   labs(x = "Hind Foot Length (mm)", y = "Weight (g)", caption = "Relationship between juvenile weight & hind foot length") + # graph labels etc
   theme(plot.caption = element_text(hjust = 0, size = 12)) #move caption to bottom left 
```

```{r}
lin_model <- lm(hindft ~ weight, data = hind_foot_hares) #linear model with hindfoot and weight
lin_model_tidy <- broom::tidy(lin_model) # creates a table
hind_int <- lin_model_tidy$estimate[1] #intercept
hind_coeff <- lin_model_tidy$estimate[2] #coefficient
hin_stats <- broom::glance(lin_model) #glance
```

